\textbf{Advection versus diffusion in brain intraventricular transport}

### 2025-03-27 paper/main.tex
update appendix

To derive weak formulations of the continuous problem defined by the Stokes
equations~\Cref{eq:appendix_stokes_eqs}, we consider the function space
\begin{align*}
    \VV &= \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gp\bigg\},
\end{align*}
for the velocity and $Q=L^2(\Omega)$ for the pressure.
Here $\Gp$ is the anterior/posterior boundary where we impose a normal pressure
boundary condition for the model versions that consider fluid motion induced by
cardiac pulsations. Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.

To derive a weak formulation of the Stokes equations~\Cref{eq:appendix_stokes_eqs},
we begin by multiplying~\Cref{eq:BDM_stokes_mom} with a
test function $\vv\in\VV$ and integrating over the domain:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv} - \intO{(\nabla\cdot\bsig)\cdot\vv} = \mathbf{0}.
\end{equation*}
Performing integration by parts of the stress term and applying Green's first identity, we get
\begin{equation}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    - \intG{\bsig\nn\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \mathbf{0}, \label{eq:modelA_weak_form1}
\end{equation}
where the symbol : signifies the matrix inner product,
$A:B=\text{trace}{(A^TB)}$ for matrices $A, B\in\mathbb{R}^{d\times d}$.

@@

We choose $\rr = \rr^+ = (1, 0, 1)$ or $\rr = \rr^- = -(1, 0, 1)$
to impose forces perpendicular to the $y$-axis.
The reason for this is that all of the experimental images we compare
with are taken in the $xz$-plane,
rendering flow features in the $y$-direction unobservable.
In the following, we consider the way of handling the stress
boundary integral for each model version successively.

\textbf{Cilia-driven/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\Cref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs,
        \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gc\cup\Gs$. The boundary $\Gs$ is a free-slip boundary,
where we impose the free-slip condition~\Cref{eq:modelA_full_slip_BC}.
The first condition~\Cref{eq:modelA_slip_BC_impermeability} is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\Cref{eq:modelA_slip_BC_shear} imposes a traction tangential to
the boundary. This tangential traction is applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the
boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar.
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition~\Cref{eq:modelA_slip_BC_shear} and the
free-slip condition~\Cref{eq:modelA_full_slip_BC}, yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
Since $\btau$ is known, the second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form.
With $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
This term vanishes owing to the construction $\VV$.
As a result, for the cilia-driven/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation~\Cref{eq:BDM_stokes_div}
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q.
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, for any real constant $C$ the vector
$(\uu, p)=(\mathbf{0}, C)$ satisfies the equations.
Hence the system~\Cref{eq:stokes_weak_modelA} is singular with a one-dimensional nullspace
of constant pressures.
Consequently, assembly of our discretization of the  equations~\Cref{eq:stokes_weak_modelA}
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

@@

\textbf{Cardiac-induced/no-cilia model.} For this model version,
the only physical mechanism inducing fluid flow is the cardiac cycle.
The cardiac cycle generates directional back-and-forth motion of the CSF
in the zebrafish brain ventricles.
We model the influence of the cardiac cycle with a normal pressure boundary condition:
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \hat{\bsig}_{\parallel} &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,
      \label{eq:modelB_slip_BC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp,
      \label{eq:modelB_pressure_BC}
    \end{alignat}%
\end{subequations}%
where the boundary $\Gamma = \Gs\cup\Gp$ consists of a free-slip boundary $\Gs$,
and a part $\Gp$ where a normal traction type of boundary condition is imposed.
Note, however, that the transpose velocity gradient term must be
included inside the parentheses on the left-hand side of~\Cref{eq:modelB_pressure_BC}
for it to be the traction $\bsig$. Since this is not the case,
we rather refer to this boundary condition as a normal pressure boundary condition.
The boundary $\Gs$ is handled as discussed in the previous section,
such that the stress boundary integral~\Cref{eq:boundary_integral_BC1}
vanishes over $\Gs$. Therefore, we now have
\begin{align}
    \intG{\hat{\bsig}\cdot\vv} = \intGp{\hat{\bsig}\cdot\vv}
    &= \intGp{(\mu(\nabla\uu + (\nabla\uu)^T) - p\mathbf{I})\nn\cdot\vv}\notag \\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{(\mu\nabla\uu
    - p\mathbf{I})\nn\cdot\vv}\notag\\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}.
    \label{eq:modelB_boundary_integral1}
\end{align}
We leave the first term on the left-hand side
of~\Cref{eq:modelB_boundary_integral1} in the weak form of
the momentum equation. The second term is determined by
the boundary condition~\Cref{eq:modelB_pressure_BC},
and is therefore moved to the right-hand side of the weak form.
The weak form of the momentum equation is then
\begin{equation}
    \begin{aligned}
    \intO{\rho\pdifft{\uu}\cdot\vv} + \intO{2\mu\beps(\uu) : \beps(\vv)} -
    \intO{p\nabla\cdot\vv} &-\intGp{\mu(\nabla\uu)^T\nn\cdot\vv} \\
    &=\intGp{\Tilde{p}(t)(\vv\cdot\nn)}, \quad\forall\vv\in\VV.
    \end{aligned}
    \label{eq:weak_form_mom_eq_modelB}
\end{equation}
Also considering the weak form of the continuity equation,
the weak formulation for the cardiac-only/no-cilia model version
then reads: find $(\uu,\, p)\in\VV\times Q$
such that~\Cref{eq:weak_form_mom_eq_modelB} and~\Cref{eq:pressure_weak_form}
hold for all $(\vv, \, q)\in\VV\times Q$.
To simulate the back-and-forth motion induced by the cardiac cycle observed in
previous work~\cite{Olstad2019CiliaryDevelopment}, we apply sinusoidal forcing.
The boundary $\Gp$ has two separate regions: one located on the
anterior ventricle and one on the posterior ventricle~(\Crefpanels{fig:fig1}{d}, blue markers).
For the normal pressure boundary condition, we set $\Tilde{p}(t) = 0$ on the anterior part and
$\Tilde{p}(t) = p_c(t) = -A\sin{\omega t}$ on the posterior part. The value of the amplitude
$A$ was based on heuristics and $\omega = 2\pi f$, where $f=2.22$ Hz
is the cardiac cycle frequency.

\textbf{Baseline model.} The baseline flow model
considers both cilia and cardiac beating as driving mechanisms of fluid flow.
We therefore consider the boundary conditions
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \bsigpar
      &= \btau  \quad &&\mathrm{on} \ \Gc, \label{eq:slip_BC_modelC} \\
      \uu\cdot\nn = 0, \ \bsigpar
      &= \mathbf{0}  \quad &&\mathrm{on} \ \Gs, \label{eq:freeslip_BC_modelC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn
      &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp, \label{eq:pressure_BC_modelC}
    \end{alignat}
\end{subequations}%
where $\Gamma = \Gc\cup\Gs\cup\Gp$ is the boundary of $\Omega$.
The quantities $\btau$ and $\Tilde{p}(t)$ have been introduced in
the previous sections covering the other flow model versions. We handle the boundary
conditions~\Cref{eq:slip_BC_modelC} and~\Cref{eq:freeslip_BC_modelC}
as introduced for the cilia-only/no-cardiac model, and the boundary
condition~\Cref{eq:pressure_BC_modelC} as introduced for the cardiac-only/no-cilia model.
The abstract weak formulation for 
the baseline model is then: find $(\uu,\, p)\in\VV\times Q$ such that
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu, \vv) + \mathcal{B}(\vv, p) &= \mathcal{L}(\vv),
        &&\quad\forall\vv\in\VV, \\
        \mathcal{B}(\uu, q) &= 0, &&\quad\forall q\in Q,
    \end{alignat}%
    \label{eq:abstract_weak_form_modelC}
\end{subequations}%
with the definitions
\begin{align}
    \mathcal{A}(\uu, \vv) &= \intO{\rho\pdifft{\uu}\cdot\vv}
                          + \intO{2\mu\beps(\uu) : \beps(\vv)}
                          -\intGp{\mu(\nabla\uu)^T\nn\cdot\vv},\label{eq:stokes_bilinear_form}\\
    \mathcal{B}(\uu, q) &= -\intO{q\nabla\cdot\uu}, \label{eq:stokes_skew_form}\\
    \mathcal{L}(\vv) &= \intGc{\btau\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}.
    \label{eq:stokes_linear_form}
\end{align}

\subsection{Discretization of the Stokes equations}
We will outline the discretization of the weak
formulation~\Cref{eq:abstract_weak_form_modelC,
eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}
of the Stokes equations for the baseline model. A similar approach can
be used to discretize the weak formulation for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. For all cells $K$ of the computational mesh $\Omega_h$,
we define
\begin{align*}
    \VV_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k(K)\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}(K); \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh.
Because $\VV_h\not\subset\VV$, the elements are non-conforming.
This BDM-DG scheme for the Stokes equations
is $\mathbf{H}(\mathrm{div})$ conforming and thus exactly mass-conserving,
because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.

>>>>

To derive weak formulations of the continuous problem defined by the Stokes
equations (\Cref{eq:appendix_stokes_eqs}), we consider the function space
\begin{align*}
    \VV &= \bigg\{\vv \ : \ \vv\in\mathbf{H}^1(\Omega),
    \vv\cdot\nn=0 \ \mathrm{on} \ \Gamma\setminus\Gp\bigg\},
\end{align*}
for the velocity and $Q=L^2(\Omega)$ for the pressure.
Here $\Gp$ is the anterior/posterior boundary where we impose a normal pressure
boundary condition for the model versions that consider fluid motion induced by
cardiac pulsations. Bold-faced characters for a space denotes a $d$--dimensional vector space,
where $d$ is the spatial dimension of $\Omega$. For example,
$\mathbf{H}^1(\Omega) = [H^1(\Omega)]^3$ if $\Omega$ is three-dimensional.

To derive a weak formulation of the Stokes equations (\Cref{eq:appendix_stokes_eqs}),
we begin by multiplying~\Cref{eq:BDM_stokes_mom} with a
test function $\vv\in\VV$ and integrating over the domain:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv} - \intO{(\nabla\cdot\bsig)\cdot\vv} = \mathbf{0}.
\end{equation*}
Performing integration by parts of the stress term and applying Green's first identity, we get
\begin{equation}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    - \intG{\bsig\nn\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \mathbf{0}, \label{eq:modelA_weak_form1}
\end{equation}
where the symbol : signifies the matrix inner product,
$A:B=\text{trace}{(A^TB)}$ for matrices $A, B\in\mathbb{R}^{d\times d}$.

@@

We choose $\rr = \rr^+ = (1, 0, 1)$ or $\rr = \rr^- = -(1, 0, 1)$
to impose forces perpendicular to the $y$-axis.
The reason for this is that all of the experimental images we compare
with are taken in the $xz$-plane (rostrocaudal-dorsoventral plane),
rendering flow features in the $y$-direction (lateral) unobservable.
In the following, we consider the way of handling the stress
boundary integral for each model version successively.

\textbf{Cilia-driven/no-cardiac model.} This model version has cilia forces
as the only mechanism driving flow. We consider the
Stokes equations~\Cref{eq:appendix_stokes_eqs} with the slip boundary conditions
\begin{subequations}
    \begin{alignat}{2}
        \uu\cdot\nn &= 0 \quad &&\mathrm{on} \ \Gamma=\Gc\cup\Gs,
        \label{eq:modelA_slip_BC_impermeability} \\
        \bsigpar &= \btau \quad &&\mathrm{on} \ \Gc,
        \label{eq:modelA_slip_BC_shear} \\
        \bsigpar &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,\label{eq:modelA_full_slip_BC}
    \end{alignat}\label{eq:modelA_slip_BC}%
\end{subequations}%
on the boundary $\Gamma=\Gc\cup\Gs$. The boundary $\Gs$ is a free-slip boundary,
where we impose the free-slip condition given by \Cref{eq:modelA_full_slip_BC}.
The first condition (\Cref{eq:modelA_slip_BC_impermeability}) is the
impermeability condition, enforcing no flow normal to the boundary,
whereas~\Cref{eq:modelA_slip_BC_shear} imposes a traction tangential to
the boundary. This tangential traction is applied to represent
the forces that the collective cilia motion exerts on the cerebrospinal fluid.

The boundary conditions defined by~\Cref{eq:modelA_slip_BC}
are set weakly through the boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form of the Stokes equations (\Cref{eq:modelA_weak_form1}).
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar.
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition (\Cref{eq:modelA_slip_BC_shear}) and the
free-slip condition (\Cref{eq:modelA_full_slip_BC}), yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
Since $\btau$ is known, the second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form.
With $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
This term vanishes owing to the construction $\VV$.
As a result, for the cilia-driven/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation (\Cref{eq:BDM_stokes_div})
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q.
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, for any real constant $C$ the vector
$(\uu, p)=(\mathbf{0}, C)$ satisfies the equations.
Hence \Cref{eq:stokes_weak_modelA} is a singular system 
with a one-dimensional nullspace of constant pressures.
Consequently, assembly of the discretization of the Stokes equations (\Cref{eq:stokes_weak_modelA})
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

@@

\textbf{Cardiac-induced/no-cilia model.} For this model version,
the only physical mechanism inducing fluid flow is the cardiac cycle.
The cardiac cycle generates directional back-and-forth motion of the CSF
in the zebrafish brain ventricles.
We model the influence of the cardiac cycle with a normal pressure boundary condition:
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \hat{\bsig}_{\parallel} &= \mathbf{0} \quad &&\mathrm{on} \ \Gs,
      \label{eq:modelB_slip_BC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp,
      \label{eq:modelB_pressure_BC}
    \end{alignat}%
\end{subequations}%
where the boundary $\Gamma = \Gs\cup\Gp$ consists of a free-slip boundary $\Gs$,
and a part $\Gp$ where a normal traction type of boundary condition is imposed.
Note, however, that the transpose velocity gradient term must be
included inside the parentheses on the left-hand side of~\Cref{eq:modelB_pressure_BC}
for it to be the traction $\bsig$. Since this is not the case,
we rather refer to this boundary condition as a normal pressure boundary condition.
The boundary $\Gs$ is handled as discussed in the previous section,
such that the stress boundary integral in~\Cref{eq:boundary_integral_BC1}
vanishes over $\Gs$. Therefore, we now have
\begin{align}
    \intG{\hat{\bsig}\cdot\vv} = \intGp{\hat{\bsig}\cdot\vv}
    &= \intGp{(\mu(\nabla\uu + (\nabla\uu)^T) - p\mathbf{I})\nn\cdot\vv}\notag \\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{(\mu\nabla\uu
    - p\mathbf{I})\nn\cdot\vv}\notag\\
    &= \intGp{\mu(\nabla\uu)^T\nn\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}.
    \label{eq:modelB_boundary_integral1}
\end{align}
We leave the first term on the left-hand side
of~\Cref{eq:modelB_boundary_integral1} in the weak form of
the momentum equation. The second term is determined by
the boundary condition defined in~\Cref{eq:modelB_pressure_BC},
and is therefore moved to the right-hand side of the weak form.
The weak form of the momentum equation is then
\begin{equation}
    \begin{aligned}
    \intO{\rho\pdifft{\uu}\cdot\vv} + \intO{2\mu\beps(\uu) : \beps(\vv)} -
    \intO{p\nabla\cdot\vv} &-\intGp{\mu(\nabla\uu)^T\nn\cdot\vv} \\
    &=\intGp{\Tilde{p}(t)(\vv\cdot\nn)}, \quad\forall\vv\in\VV.
    \end{aligned}
    \label{eq:weak_form_mom_eq_modelB}
\end{equation}
Also considering the weak form of the continuity equation,
the weak formulation for the cardiac-only/no-cilia model version
then reads: find $(\uu,\, p)\in\VV\times Q$
such that~\Cref{eq:weak_form_mom_eq_modelB} and~\Cref{eq:pressure_weak_form}
hold for all $(\vv, \, q)\in\VV\times Q$.
To simulate the back-and-forth motion induced by the cardiac cycle observed in
previous work~\cite{Olstad2019CiliaryDevelopment}, we apply sinusoidal forcing.
The boundary $\Gp$ has two separate regions: one located on the
anterior ventricle and one on the posterior ventricle~(\Crefpanels{fig:fig1}{d}, blue markers).
For the normal pressure boundary condition, we set $\Tilde{p}(t) = 0$ on the anterior part and
$\Tilde{p}(t) = p_c(t) = -A\sin{\omega t}$ on the posterior part. The value of the amplitude
$A$ was based on heuristics and $\omega = 2\pi f$, where $f=2.22$ Hz
is the cardiac cycle frequency.

\textbf{Baseline model.} The baseline flow model
considers both cilia and cardiac beating as driving mechanisms of fluid flow.
We therefore consider the boundary conditions
\begin{subequations}
    \begin{alignat}{2}
      \uu\cdot\nn = 0, \ \bsigpar
      &= \btau  \quad &&\mathrm{on} \ \Gc, \label{eq:slip_BC_modelC} \\
      \uu\cdot\nn = 0, \ \bsigpar
      &= \mathbf{0}  \quad &&\mathrm{on} \ \Gs, \label{eq:freeslip_BC_modelC} \\
      (\mu\nabla\uu - p\mathbf{I})\nn
      &= \Tilde{p}(t)\nn \quad &&\mathrm{on} \ \Gp, \label{eq:pressure_BC_modelC}
    \end{alignat}
\end{subequations}%
where $\Gamma = \Gc\cup\Gs\cup\Gp$ is the boundary of $\Omega$.
The quantities $\btau$ and $\Tilde{p}(t)$ have been introduced in
the previous sections covering the other flow model versions. We handle the boundary
conditions~\Cref{eq:slip_BC_modelC} and~\Cref{eq:freeslip_BC_modelC}
as introduced for the cilia-only/no-cardiac model, and the boundary
condition~\Cref{eq:pressure_BC_modelC} as introduced for the cardiac-only/no-cilia model.
The abstract weak formulation for 
the baseline model is then: find $(\uu,\, p)\in\VV\times Q$ such that
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu, \vv) + \mathcal{B}(\vv, p) &= \mathcal{L}(\vv),
        &&\quad\forall\vv\in\VV, \\
        \mathcal{B}(\uu, q) &= 0, &&\quad\forall q\in Q,
    \end{alignat}%
    \label{eq:abstract_weak_form_modelC}
\end{subequations}%
with the definitions
\begin{align}
    \mathcal{A}(\uu, \vv) &= \intO{\rho\pdifft{\uu}\cdot\vv}
                          + \intO{2\mu\beps(\uu) : \beps(\vv)}
                          -\intGp{\mu(\nabla\uu)^T\nn\cdot\vv},\label{eq:stokes_bilinear_form}\\
    \mathcal{B}(\uu, q) &= -\intO{q\nabla\cdot\uu}, \label{eq:stokes_skew_form}\\
    \mathcal{L}(\vv) &= \intGc{\btau\cdot\vv} + \intGp{\Tilde{p}(t)(\vv\cdot\nn)}.\label{eq:stokes_linear_form}
\end{align}

\subsection{Discretization of the Stokes equations}
We outline the discretization of the weak
formulation of the Stokes equations for the baseline model 
(both cilia traction and cardiac pressure pulsations present) given by
\Cref{eq:abstract_weak_form_modelC,eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}.
A similar approach can be used to discretize the weak formulation
for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. For all cells $K$ of the computational mesh $\Omega_h$,
we define
\begin{align*}
    \VV_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k(K)\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}(K); \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh.
Because $\VV_h\not\subset\VV$, the elements are non-conforming.
This BDM-DG scheme for the Stokes equations
is $\mathbf{H}(\mathrm{div})$ conforming and thus exactly mass-conserving,
because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.
